Evidence for a Color Parameter Unassociated With Dust Within the Type~Ia Supernovae of the Nearby Supernova Factory

### 2016-10-11 forgerard.tex
more Greg comments

We hypothesize that at peak brightness
SN~Ia broadband magnitudes and colors are correlated with
spectral features: equivalent widths and line positions are considered as statistics localized in wavelength that are insensitive to variations in
broadband colors.
%Spectra may not efficiently capture all variations in broadband magnitude, so an additional source
%of intrinsic variation is accommodated.
These intrinsic spectral  parameters may not deterministically predict magnitudes, but rather do so with some intrinsic dispersion.
\textcolor{red}{The intrinsic magnitudes are then
modified by an extrinsic physical process (i.e.\ dust) to produce apparent magnitudes.}
\subsection{Model}
We assume 
that  peak intrinsic $UBVRI$ magnitudes are linearly dependent
on the
 equivalent widths of the CaII H\&K and SiII$\lambda$4141 spectral features
$EW_{Ca}$, $EW_{Si}$,
\textcolor{red}{ and $\lambda_{Si}$ the wavelength of the minimum of 
the SiII$\lambda6355$ feature}:
these spectral features are associated with SN~Ia  spectroscopic diversity  
\citep{2006PASP..118..560B, 2008A&A...492..535A, 2009A&A...500L..17B, 2009PASP..121..238B, 2009ApJ...699L.139W, 2011ApJ...729...55F}.
\textcolor{red}{
The explicit omission of light-curve shape in our model is compensated by its proxy
$EW_{Si}$ at peak brightness}
\citep{2008A&A...492..535A, 2011A&A...529L...4C}. 
% An intrinsic-color parameter $D$ describes  intrinsic magnitude/color variation unaccounted for by the spectral features.
Residual dispersion is described by a Normal distribution with a parameterized covariance matrix
$C_c$.  A grey magnitude offset, $\Delta$, is included for each supernova
to capture intrinsic dispersion and peculiar-velocity errors introduced when converting fluxes to luminosity.
The observed magnitudes are linearly dependent on the
extrinsic-color parameters $k_0$ and $k_1$.  The observables
$U_o, B_o, V_o, R_o, I_o$, $EW_{Ca,o}$, $EW_{Si,o}$, $\lambda_{Si,o}$
shown in Table~\ref{data:tab} have Gaussian measurement uncertainty with covariance $C$.
The model is written as
\begin{equation}
\begin{pmatrix}
U\\B\\V\\R\\I
\end{pmatrix}
\sim \mathcal{N}
\left(
\Delta +
\begin{pmatrix}
c_0+\alpha_0 EW_{Ca} + \beta_0 EW_{Si} + \eta_0 \lambda_{Si} \\
c_1+\alpha_1 EW_{Ca} + \beta_1 EW_{Si} + \eta_1 \lambda_{Si}  \\
c_2+\alpha_2 EW_{Ca} + \beta_2 EW_{Si} + \eta_2 \lambda_{Si} \\
c_3+\alpha_3 EW_{Ca} + \beta_3 EW_{Si} + \eta_3 \lambda_{Si} \\
c_4+\alpha_4 EW_{Ca} + \beta_4 EW_{Si}+ \eta_4 \lambda_{Si}
\end{pmatrix}
,C_{c}
\right)
\label{ewsiv:eqn}
\end{equation}
\begin{equation}
\begin{pmatrix}
U_o\\B_o\\ V_o\\R_o\\I_o\\EW_{Si, o}\\ EW_{Ca, o} \\ \lambda_{Si, o}
\end{pmatrix}
\sim \mathcal{N}
\left(
\begin{pmatrix}
U +\gamma^0_{1} k_0 +\gamma^1_{1} k_1 \\B +\gamma^0_{1} k_0 +\gamma^1_{1} k_1 \\
V+\gamma^0_{2} k_0+\gamma^1_{2} k_1\\R+\gamma^0_{3} k_0 + \gamma^1_{3} k_1\\I+\gamma^0_{4} k_0+\gamma^1_{4} k_1\\
EW_{Si}\\ EW_{Ca} \\ \lambda_{Si}
\end{pmatrix}
,C
\right).
\label{dust:eqn}
\end{equation}
The global parameters $c$, $\alpha$, $\beta$,  and $\eta$,  are the intercepts and slopes of the linear relationships that
relate intrinsic magnitudes with the spectral-feature parameters.
(In this article we freely interchange the parameter indices with  $01234$ or $UBVRI$.)
The global parameters $\gamma^0$, $\gamma^1$  are the slopes that connect the extrinsic-color
parameters to observed magnitudes.

@@

\section{Supernova Model II: Two Extrinsic Parameter, One Intrinsic Parameter}
\label{model2:sec}
In \S\ref{model:sec} we show that the Nearby Supernova Factory data analyzed with an agnostic model with
two extrinsic parameters produce
results consistent
with two-parameter dust models.  An extension to our model is now added to accommodate
a possible additional
color parameter.
The new model is written as
\begin{equation}
\begin{pmatrix}
U\\B\\V\\R\\I
\end{pmatrix}
\sim \mathcal{N}
\left(
\Delta +
\begin{pmatrix}
c_0+\alpha_0 EW_{Ca} + \beta_0 EW_{Si} + \eta_0 \lambda_{Si} + \delta_0 D\\
c_1+\alpha_1 EW_{Ca} + \beta_1 EW_{Si} + \eta_1 \lambda_{Si} + \delta_1 D \\
c_2+\alpha_2 EW_{Ca} + \beta_2 EW_{Si} + \eta_2 \lambda_{Si} + \delta_2 D\\
c_3+\alpha_3 EW_{Ca} + \beta_3 EW_{Si} + \eta_3 \lambda_{Si} + \delta_3 D\\
c_4+\alpha_4 EW_{Ca} + \beta_4 EW_{Si}+ \eta_4 \lambda_{Si} + \delta_4 D
\end{pmatrix}
,C_{c}
\right)
\label{ewsiv:eqn}
\end{equation}
\begin{equation}
\begin{pmatrix}
U_o\\B_o\\ V_o\\R_o\\I_o\\EW_{Si, o}\\ EW_{Ca, o} \\ \lambda_{Si, o}
\end{pmatrix}
\sim \mathcal{N}
\left(
\begin{pmatrix}
U +\gamma^0_{1} k_0 +\gamma^1_{1} k_1 \\B +\gamma^0_{1} k_0 +\gamma^1_{1} k_1 \\
V+\gamma^0_{2} k_0+\gamma^1_{2} k_1\\R+\gamma^0_{3} k_0 + \gamma^1_{3} k_1\\I+\gamma^0_{4} k_0+\gamma^1_{4} k_1\\
EW_{Si}\\ EW_{Ca} \\ \lambda_{Si}
\end{pmatrix}
,C
\right).
\label{dust:eqn}
\end{equation}
In all aspects it is identical to the previous model except for the addition of the term $\delta D$,
which affects the mean
absolute magnitudes:
each supernova has an intrinsic $D$ parameter that is linearly related to
absolute magnitudes through global coefficients $\delta$. 
Although
this term is introduced through absolute rather than observed magnitudes,
practically identical results are obtained when it is placed alongside the extrinsic $\gamma k$ terms.  Despite this
degeneracy, for convenience this term is referred to as ``intrinsic''.

To constrain the new degrees of freedom and degeneracies inherent in the model we impose that
$D+1/N$ be a simplex to break the 
degeneracy  $\delta \rightarrow a\delta$, $D \rightarrow a^{-1}D$.
Furthermore we impose
\begin{equation}
\langle D\rangle=0,  \delta \cdot \gamma^0=0, \delta \cdot \gamma^1=0.
\end{equation}
The first specifies the definition of zero color.   The latter two confine possible magnitude
variations to those that cannot be attributed to 
the $\gamma k$ terms (i.e.\ dust); this condition breaks the degeneracies between $\delta D$  
and $\gamma k$. 
Unlike for the $\gamma$'s, there is no condition on the sign of $\delta_0$,
so that the posterior has degenerate solutions with sign flips of $\delta$ and $D$.
In the running of the Monte Carlo, the initial conditions for
$\gamma^0$ and $\gamma^1$ were set to be in the neighborhood of the best fit of \S\ref{model:sec}
to maintain their correspondence between the two models.

The resulting credible intervals of our parameters are given in Table~\ref{global2:tab}.  Comparison with the
results of the 2-parameter extrinsic model given in Table~\ref{global:tab} shows that  parameters that are common to both
are consistent well within 1$\sigma$ uncertainties.  The most significant changes of note are in the standard
deviations of the residuals in the $V$ and $I$ bands.
The newly fitted $\delta$ parameters have an almost identical association with the Fitzpatrick dust law, in that now
\begin{equation}
\begin{pmatrix}
 \frac{\gamma^0_X}{\gamma^0_1-\gamma^0_2} \\
\frac{\gamma^1_X}{\gamma^1_1-\gamma^1_2} 
\end{pmatrix}=
\begin{pmatrix}
\begin{array}{rr}
1.16 & 2.82 \\
0.74 & -5.43
\end{array}
\end{pmatrix} 
\begin{pmatrix}
a(X) \\
b(X)
\end{pmatrix}+
\vec{\epsilon},
\end{equation}
where the residual $\vec{\epsilon}$''s perpendicular to the Fitzpatrick plane
contribute similar 2\% and 5\% to the total  length of the vector.

@@

The distributions of both sources of color excess $E(B-V)$ and
 $E_\delta(B-V) = (\delta_B-\delta_V)D$ are shown in the ideograms in Figure~\ref{ebv:fig}.
The standard deviations of $E_\gamma(B-V)$ and $E_\delta(B-V)$ are
%-----
0.081
and 0.010
%-----
mag respectively.
The range of $B-V$ colors is $\sim 8$ times larger for the external contribution; variation in $B-V$ color due to the intrinsic term is dwarfed by the extrinsic process. 
Both distributions are non-Gaussian, that of $E(B-V)$ having a sharp rise in the blue and an extended tail in the red, consistent
with the expectation for extinction due to host-galaxy dust \citep{1998ApJ...502..177H}.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.8in]{output15/ebv.pdf}
   \includegraphics[width=2.8in]{output15/ebv_delta.pdf}
      \caption{Left: Ideograms of the external $E(B-V)$ and
   internal $E_\delta(B-V) = (\delta_B-\delta_V)D$  contributions to color excess  for the supernovae in our sample.
   Right: Ideogram of $E_\delta(B-V)$ on an enlarged scale.
   \label{ebv:fig}}
\end{figure}
The significance of the $\delta$ parameters is shown pictorially
in Figure~\ref{deltacorner:fig}. The left plot shows the $\delta$ posteriors, which are bimodal due to the sign-flip degeneracy.  The probability
of the null result $\delta_X=0$ for all $X$ is $<98\%$, determined through
the population of its bin in a 5-dimensional histogram.
The right plot shows values of $\delta_X/\delta_I-1$ that  have significant non-zero values.   An interesting feature
is that the shift in magnitude is not monotonic in wavelength, but rather decreases from $U$ to $V$, and then increases from $V$
to $I$.
These are not the behaviors expected from normal dust,
circumstellar dust suggested as a source of SN extinction \citep{2005ApJ...635L..33W,2008ApJ...686L.103G,
2015ApJ...807L..26G}, or a blackbody in the Raleigh-Jeans tail, and so may be indicative of
more complicated underlying supernova physics.

>>>>

We hypothesize that at peak brightness
SN~Ia broadband magnitudes and colors are correlated with
spectral features: equivalent widths and line positions are considered as statistics localized in wavelength that are insensitive to variations in
broadband colors.
%Spectra may not efficiently capture all variations in broadband magnitude, so an additional source
%of intrinsic variation is accommodated.
These intrinsic spectral  parameters may not deterministically predict magnitudes, but rather do so with some intrinsic dispersion.
\textcolor{red}{The intrinsic magnitudes are then
modified by an extrinsic physical process (i.e.\ dust) to produce apparent magnitudes.}
\subsection{Model}
We assume 
that  peak intrinsic $UBVRI$ magnitudes are linearly dependent
on the
 equivalent widths of the CaII H\&K and SiII$\lambda$4141 spectral features
$EW_{Ca}$, $EW_{Si}$,
\textcolor{red}{ and $\lambda_{Si}$ the wavelength of the minimum of 
the SiII$\lambda6355$ feature}:
these spectral features are associated with SN~Ia  spectroscopic diversity  
\citep{2006PASP..118..560B, 2008A&A...492..535A, 2009A&A...500L..17B, 2009PASP..121..238B, 2009ApJ...699L.139W, 2011ApJ...729...55F}.
\textcolor{red}{
The explicit omission of light-curve shape in our model is compensated by its proxy
$EW_{Si}$ at peak brightness}
\citep{2008A&A...492..535A, 2011A&A...529L...4C}. 
% An intrinsic-color parameter $D$ describes  intrinsic magnitude/color variation unaccounted for by the spectral features.
Residual dispersion is described by a Normal distribution with a parameterized covariance matrix
$C_c$.  A grey magnitude offset, $\Delta$, is included for each supernova
to capture intrinsic dispersion and peculiar-velocity errors introduced when converting fluxes to luminosity.
The observed magnitudes are linearly dependent on the
extrinsic-color parameters $k_0$ and $k_1$.  
The observables
$U_o, B_o, V_o, R_o, I_o$, $EW_{Ca,o}$, $EW_{Si,o}$, $\lambda_{Si,o}$
shown in Table~\ref{data:tab} have Gaussian measurement uncertainty with covariance $C$.
\textcolor{red}{Unlike the parameters associated with
the spectral measurements, $EW_{Ca}$, $EW_{Si}$ and $\lambda_{Si}$,  $k_0$ and $k_1$ are latent parameters with no
directly associated observables.}
The model is written as
\begin{equation}
\begin{pmatrix}
U\\B\\V\\R\\I
\end{pmatrix}
\sim \mathcal{N}
\left(
\Delta +
\begin{pmatrix}
c_0+\alpha_0 EW_{Ca} + \beta_0 EW_{Si} + \eta_0 \lambda_{Si} \\
c_1+\alpha_1 EW_{Ca} + \beta_1 EW_{Si} + \eta_1 \lambda_{Si}  \\
c_2+\alpha_2 EW_{Ca} + \beta_2 EW_{Si} + \eta_2 \lambda_{Si} \\
c_3+\alpha_3 EW_{Ca} + \beta_3 EW_{Si} + \eta_3 \lambda_{Si} \\
c_4+\alpha_4 EW_{Ca} + \beta_4 EW_{Si}+ \eta_4 \lambda_{Si}
\end{pmatrix}
,C_{c}
\right)
\label{ewsiv:eqn}
\end{equation}
\begin{equation}
\begin{pmatrix}
U_o\\B_o\\ V_o\\R_o\\I_o\\EW_{Si, o}\\ EW_{Ca, o} \\ \lambda_{Si, o}
\end{pmatrix}
\sim \mathcal{N}
\left(
\begin{pmatrix}
U +\gamma^0_{1} k_0 +\gamma^1_{1} k_1 \\B +\gamma^0_{1} k_0 +\gamma^1_{1} k_1 \\
V+\gamma^0_{2} k_0+\gamma^1_{2} k_1\\R+\gamma^0_{3} k_0 + \gamma^1_{3} k_1\\I+\gamma^0_{4} k_0+\gamma^1_{4} k_1\\
EW_{Si}\\ EW_{Ca} \\ \lambda_{Si}
\end{pmatrix}
,C
\right).
\label{dust:eqn}
\end{equation}
The global parameters $c$, $\alpha$, $\beta$,  and $\eta$,  are the intercepts and slopes of the linear relationships that
relate intrinsic magnitudes with the spectral-feature parameters.
(In this article we freely interchange the parameter indices with  $01234$ or $UBVRI$.)
The global parameters $\gamma^0$, $\gamma^1$  are the slopes that connect the extrinsic-color
parameters to observed magnitudes.

@@

\section{Supernova Model II: Two Extrinsic Parameter, One Intrinsic Parameter}
\label{model2:sec}
In \S\ref{model:sec} we show that the Nearby Supernova Factory data analyzed with an agnostic model with
two extrinsic parameters produce
magnitudes consistent
with two-parameter dust models that have been established for stars in the Milky Way, LMC, and SMC.
An extension to our model is now added to explore the possibility of
a possible additional
color parameter that could have its origins in supernova astrophysics.
The extended model is written as
\begin{equation}
\begin{pmatrix}
U\\B\\V\\R\\I
\end{pmatrix}
\sim \mathcal{N}
\left(
\Delta +
\begin{pmatrix}
c_0+\alpha_0 EW_{Ca} + \beta_0 EW_{Si} + \eta_0 \lambda_{Si} + \delta_0 D\\
c_1+\alpha_1 EW_{Ca} + \beta_1 EW_{Si} + \eta_1 \lambda_{Si} + \delta_1 D \\
c_2+\alpha_2 EW_{Ca} + \beta_2 EW_{Si} + \eta_2 \lambda_{Si} + \delta_2 D\\
c_3+\alpha_3 EW_{Ca} + \beta_3 EW_{Si} + \eta_3 \lambda_{Si} + \delta_3 D\\
c_4+\alpha_4 EW_{Ca} + \beta_4 EW_{Si}+ \eta_4 \lambda_{Si} + \delta_4 D
\end{pmatrix}
,C_{c}
\right)
\label{ewsiv:eqn}
\end{equation}
\begin{equation}
\begin{pmatrix}
U_o\\B_o\\ V_o\\R_o\\I_o\\EW_{Si, o}\\ EW_{Ca, o} \\ \lambda_{Si, o}
\end{pmatrix}
\sim \mathcal{N}
\left(
\begin{pmatrix}
U +\gamma^0_{1} k_0 +\gamma^1_{1} k_1 \\B +\gamma^0_{1} k_0 +\gamma^1_{1} k_1 \\
V+\gamma^0_{2} k_0+\gamma^1_{2} k_1\\R+\gamma^0_{3} k_0 + \gamma^1_{3} k_1\\I+\gamma^0_{4} k_0+\gamma^1_{4} k_1\\
EW_{Si}\\ EW_{Ca} \\ \lambda_{Si}
\end{pmatrix}
,C
\right).
\label{dust:eqn}
\end{equation}
In all aspects it is identical to the previous model except for the addition of the term $\delta D$,
which affects the mean
absolute magnitudes:
each supernova has an intrinsic $D$ parameter that is linearly related to
absolute magnitudes through global coefficients $\delta$.  Note that $D$ is a latent supernova characteristic whose existence is inferred from
its effects on broadband magnitudes.
Although
this term is introduced through absolute rather than observed magnitudes,
practically identical results are obtained when it is placed alongside the extrinsic $\gamma k$ terms.  Despite this
degeneracy, for convenience this term is referred to as ``intrinsic''.

To constrain the new degrees of freedom and degeneracies inherent in the model we impose that
$D+1/N$ be a simplex to break the 
degeneracy  $\delta \rightarrow a\delta$, $D \rightarrow a^{-1}D$.
Furthermore we impose
\begin{equation}
\langle D\rangle=0,  \delta \cdot \gamma^0=0, \delta \cdot \gamma^1=0.
\end{equation}
The first constraint specifies the definition of zero color.   The latter two confine possible magnitude
variations to those that cannot be attributed to 
the $\gamma k$ terms (i.e.\ dust); this condition breaks the degeneracies between $\delta D$  
and $\gamma k$. 
Unlike for the $\gamma$'s, there is no condition on the sign of $\delta_0$,
so that the posterior has degenerate solutions with sign flips of $\delta$ and $D$.
In the running of the Monte Carlo, the initial conditions for
$\gamma^0$ and $\gamma^1$ were set to be in the neighborhood of the best fit of \S\ref{model:sec}
to maintain their correspondence between the two models.

The resulting credible intervals of our parameters are given in Table~\ref{global2:tab}.  Comparison with the
results of the 2-parameter extrinsic model given in Table~\ref{global:tab} shows that  parameters that are common to both
are consistent well within 1$\sigma$ uncertainties.  The most significant changes of note are in the standard
deviations of the residuals in the $V$ and $I$ bands, which are now reduced.
The newly fitted $\gamma$ parameters have an almost identical association with the Fitzpatrick dust law, in that now
\begin{equation}
\begin{pmatrix}
 \frac{\gamma^0_X}{\gamma^0_1-\gamma^0_2} \\
\frac{\gamma^1_X}{\gamma^1_1-\gamma^1_2} 
\end{pmatrix}=
\begin{pmatrix}
\begin{array}{rr}
1.16 & 2.82 \\
0.74 & -5.43
\end{array}
\end{pmatrix} 
\begin{pmatrix}
a(X) \\
b(X)
\end{pmatrix}+
\vec{\epsilon},
\end{equation}
where the residual vectors $\vec{\epsilon}$ are perpendicular to the Fitzpatrick plane and
contribute similar 2\% and 5\% to the total  length of the vector.

@@

The distributions of both sources of color excess $E(B-V)$ and
 $E_\delta(B-V) = (\delta_B-\delta_V)D$ are shown in the ideograms in Figure~\ref{ebv:fig}.
The standard deviations of $E_\gamma(B-V)$ and $E_\delta(B-V)$ are
%-----
0.081
and 0.010
%-----
mag respectively.
The range of $B-V$ colors is $\sim 8$ times larger for the external contribution; variation in $B-V$ color due to the intrinsic term is dwarfed by the extrinsic dust
component. 
Both distributions are non-Gaussian, that of $E(B-V)$ having a sharp rise in the blue and an extended tail in the red, consistent
with the expectation of dust considering the distribution of supernovae within galaxies and the distribution of galaxy orientations with respect to the observer \citep{1998ApJ...502..177H}.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.8in]{output15/ebv.pdf}
   \includegraphics[width=2.8in]{output15/ebv_delta.pdf}
      \caption{Left: Ideograms of the external $E(B-V)$ and
   internal $E_\delta(B-V) = (\delta_B-\delta_V)D$  contributions to color excess  for the supernovae in our sample.
   Right: Ideogram of $E_\delta(B-V)$ on an enlarged scale.
   \label{ebv:fig}}
\end{figure}
The significance of the $\delta$ parameters is shown pictorially
in Figure~\ref{deltacorner:fig}. The left plot shows the $\delta$ posteriors, which are bimodal due to the sign-flip degeneracy.  The probability
of the null result $\delta_X=0$ for all $X$ is $<98\%$, as determined through
the population of its bin in a 5-dimensional histogram.
The right plot shows values of $\delta_X/\delta_I-1$ that  have significant non-zero values.   An interesting feature
is that the shift in magnitude is not monotonic in wavelength, but rather decreases from $U$ to $V$, and then increases from $V$
to $I$.
These are not the behaviors expected from normal dust,
circumstellar dust suggested as a source of SN extinction \citep{2005ApJ...635L..33W,2008ApJ...686L.103G,
2015ApJ...807L..26G} and so may be indicative of underlying supernova physics beyond that captured by $EW_{Ca}$, $EW_{Si}$ and
$\lambda_{Si}$.
